Adding Local Weather Simulation to the BigPetStore Data Generator for the JVM

### 2014-12-29 bdse2015_paper/paper.tex
updates

% *** SUBFIGURE PACKAGES ***
\usepackage[tight,footnotesize]{subfigure}
% subfigure.sty was written by Steven Douglas Cochran. This package makes it
% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
% work, it is a good idea to load it with the tight package option to reduce
% the amount of white space around the subfigures. subfigure.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at:
% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
% subfigure.sty has been superceeded by subfig.sty.

@@

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Adding Local Weather Simulation to the BigPetStore Data Generator}

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Ronald J. Nowling}
\IEEEauthorblockA{Red Hat, Inc.\\
Raleigh, NC 27601\\
Computer Science \& Engineering,\\
University of Notre Dame,\\
Notre Dame, IN 46617\\
Email: rnowling@redhat.com}
\and
\IEEEauthorblockN{Jay Vyas}
\IEEEauthorblockA{Red Hat, Inc.\\
Raleigh, NC 27601\\
Email: jvyas@redhat.com}}

@@

\section{Overview of Model \& Simulation Procedures}
Our generative model combines various well-known mathematical modeling techniques to simulate the factors affecting customers' purchasing habits.  Each part of the model is derived from \emph{ab initio} assumptions.  In several cases, real data is used to parameterize parts of the model.

\subsection{Description of Generated Data}
The model generates data for four types of records: stores, customers, transactions, and transaction items (Figure~\ref{fig:relational-data-model}).  Store records contain both a unique identifier and a location in the form of a 5-digit zip code. Customer records consist of a unique identifier, a name, and a location given as a 5-digit zip code. Transaction records consist of a transaction identifier, a customer identifier, a store identifier, and a time given in days since the beginning of the simulation. Transaction item records contain a transaction identifier and an item description.  The item description is a list of key-value pairs stored as a string.  Key-value pairs are used since item characteristics differ depending on the type of item.  For example, dog food has a brand, a flavor, a size (in lbs), and a per-unit cost, while poop bags have a brand, a color, a count, and a per-unit cost.

@@

\subsection{Generation of Stores}
The locations of the stores are modeled by a probability density function (PDF) that gives the probability that a store's location is the given zip code. We designed the PDF to give to zip codes in high-population, high-income areas the highest probabilities. The PDF is composed of two individual PDFs. One PDF determines the probability of each zip code as the population of that zip code over the total population:

\begin{equation*}
p(\text{location}=z | \text{population}(z)) = \frac{\text{population}(z)}{\sum_{i} \text{population}(i)}
\end{equation*}

The second PDF scales the probabilities of the zip codes by their incomes.  The zip code with the highest-income has a probability $s$-times larger than that of the lowest-income zip code. The values in-between are interpolated using an exponential function:

\begin{equation*}
p(\text{location}=z | \text{income}(z)) = s ^ {\big( \frac{\text{income} - \min_i{\textrm{income}(i)}} {\max_i{\textrm{income}(i)} - \min_i{\textrm{income}(i)}} - 1 \big)}
\end{equation*}

The combined PDF is given as:

\begin{align*}
p(\text{location}=z) = &Z p(\text{location}=z | \text{population}(z)) \\
&p(\text{location}=z | \text{income}(z))
\end{align*}

where $Z$ is the normalization factor.  Given the small size ($\approx$30,000 zip codes) of the data set, $Z$ can be found directly by iterating over all of the zip codes and summing the scores. The population and household income data for the zip codes were taken from from the U.S. Census American Community Survey \cite{ACS}.  The stores' locations are generated by sampling zip codes from the PDF.

\subsection{Generation of Customers}
For each proposed customer location zip code $z$, we compute the distance $d_m(z)$ between $z$ and each store's zip code $s_z$ to find the closest store.  The zip codes' latitudes and longitudes (taken from the Zip Code Database Project \cite{Zips}) are used to compute the distances. Each zip code $z$ is assigned a weight $w_z$ according to its distance $d_m(z)$ to the nearest store using an exponential distribution with the average distance $\beta$:

\begin{align*}
&w_z = \beta^{-1} \exp(-\beta^{-1} \, d_m(z)) \\
&d_m(z) = \min_{s_z} \, d(z, s_z)
\end{align*}

The customer's zip code is chosen by sampling from the zip codes with the probabiltiy of choosing each zip code $z$ proportional to its weight $w_z$.

Names are generated using data from the Name Database\cite{NameDB}. Each record in the database gives a name, a weight, and flag indicatings if the name can be used as a first name, a last name, or both.  PDFs generated for the first and last names using the weights.  The customer's name is generated by sampling a first name and a last name from each PDF respectively.

We determine the number of pets $N_p$ each customer has by sampling from a discrete uniform distribution of integers. We then sample the number of dogs $N_d$ as a discrete uniform distribution of integers between 0 and $N_p$.  The remaining number of pets are assigned to be cats.

\begin{align*}
N_c = N_p - N_d \\
N_d \sim U(0, N_p) \\
N_p \sim U(1, b)
\end{align*}

\subsection{Simulation of Transactions}
A transaction simulation is run for each customer. The transaction's store is set to the store located closest to the customer.  Transaction times are generated from a Monte Carlo process that proposes transaction times based on the usage of the customer's items and Poisson processes modeling the amount of time between the customer visiting the store and running out of the items (Section~\ref{sec:transaction-times}).  The purchased items are generated by a Hidden Markov Model (HMM) parameterized by the transaction time and amount of time remaining for the items in the customer's inventory (Section~\ref{sec:transaction-purchases}).

\begin{figure}[!t]
  \centering
  \includegraphics[width=3.5in]{figures/transaction_simulation.eps}
  \caption{Flowchart of Transaction Monte Carlo Simulation}
  \label{fig:trans_sim}
\end{figure}

\subsubsection{Simulation of Transaction Times} \label{sec:transaction-times}
Transaction times are simulated using a Monte Carlo method (Figure~\ref{fig:trans_sim}).  The usage over time of each item category is simulated. Items categories are groups of items which are interchangeable such as ``dry dog food,'' ``dry cat food,'' and ``kitty litter.'' The time between the customer's visit to the store to buy more items in each category and the exhaustion time of that item category is modeled using a Poisson process. Proposed transaction times for each item category are computed based on the exhaustion time and time sampled from the Poisson process. The earliest transaction time is taken as the overall proposed transaction time.  The probability of the transaction time is calculated using a PDF.  If rejected, new transaction times are proposed.  Otherwise, the purchased items are modeled using a separate process (discussed below). After the items are chosen, the process begins again with the simulation of the item category usages.

In the first stage of the process, the usage of items over time is modeled using a stochastic differential equation (SDE):

\begin{equation*}
\frac{da_i}{dt} = \min \{-\mu_i + \sigma_i\, dW_i(t), 0\}
\end{equation*}

where $a_i$ is the amount of item category $i$ ``stuff'' remaining at time $t$. The variable $\mu_i$ gives the average usage rate and $\sigma^2_i$ gives the variance of the usage rate for item category $i$. The SDE is numerically integrated using the Euler-Maruyama method\cite{Klouden13} with time step $\Delta t_n$ sampled from an exponential distribution:

\begin{align*}
&a_{n+1, i} = a_{n,i} - \min \{\mu_{i, c} \, \Delta t_n + \sigma_{i, c} \, \sqrt{\Delta t_n} R_n, 0.0\} \\
&t_{n+1} = t_n + \Delta t_n \\
&\Delta t_n \sim \text{Exp}(\beta^{-1}_{i,c}) \\
&R_n \sim \text{N}(0, 1)
\end{align*}

where $a_{n, i}$ is the amount remaining of item category $i$ at time step $t_n$, $\mu_{i,c}$ is the average amount used per time used, $\sigma^2_{i,c}$ is the variance of the amount used per time used, and $\beta_{i,c}$ is the item category's average amount of time between uses for customer $c$.  The parameters $\mu_{i, c}$, and $\sigma^2_{i,c}$ are computed from a base rate for each item category multiplied by the number of pets of the appropriate species the customer $c$ has.  The exhaustion time $T_{E,i}$ for each item category is found by simulating the usage until $a_{n,i} \leq 0$.

The proposed transaction time is found from the exhaustion times (Eq.~\ref{eq:transaction-times}).  For each item category, the offset time $T_{O, i}$ between when a customer would visit the store and the exhaustion time is sampled from an exponential distribution. The distribution is parameterized by $\beta_c$, the average number of days before triggering a transaction. The parameter $\beta_c$ is set separately for each customer $c$ by sampling from a uniform distribution. A proposed transaction time $T_{P, i}$ for each item category is found by subtracting the offset time $T_{O, i}$ from the exhaustion time $T_{E, i}$.  The earliest proposed transaction time is taken as the overall proposed transaction time $T_T$.

\begin{align} \label{eq:transaction-times}
&T_T = \min_i \, \{  T_{P, i}\} \\
&T_{P, i} = T_{E,i} - T_{O, i} \nonumber\\
&T_{O, i} \sim \text{Exp}(\beta^{-1}_c) \nonumber \\
&\beta_c \sim \text{U}(a, b) \nonumber
\end{align}

The acceptance probability $p(t_{n+1}=T_T|t_n)$ of the proposed transaction time $T_T$ is modeled using a PDF (Eq.~\ref{eq:transaction_time_pdf}). For now, the PDF only ensures that the proposed transaction time is more recent than the last transaction time.

%The PDF could be extended to incorporate additional factors including time-dependent events such as sales, weather, and the amount of money the customer has available at the time.

\begin{equation} \label{eq:transaction_time_pdf}
p(t_{n+1}=T_T|t_n) = \left\{ 
  \begin{array}{l l}
   1 & \quad t_{n+1} \geq t_n\\
   0 & \quad t_{n+1} < t_n
  \end{array} \right.
\end{equation}

The proposed transaction time is accepted if $p(t_{n+1}=T_T) < r$, where $r \sim \text{U}(0, 1)$. Until the proposed transaction time is accepted, a new proposed transaction time is generated by sampling new offset times. If the proposed transaction time is accepted, the purchased items are chosen through a separate simulation discussed in Section~\ref{sec:transaction-purchases} below. Transactions are generated until the accepted transaction time is later than the end time given as a simulation parameter.

\begin{figure*}[!t]
  \centering
  \includegraphics[width=6in]{figures/shopping_cart_simulation.eps}
  \caption{Example Shopping Cart Hidden Markov Model}
  \label{fig:shopping_cart_sim}
\end{figure*}

\subsubsection{Simulation of Transaction Items} \label{sec:transaction-purchases}

A customer's purchases in each transaction are modeled using a layered Hidden Markov Model (HMM) (Figure~\ref{fig:shopping_cart_sim}).  The HMM has three types of states: the start state, the purchase states, and the stop state. There are an infinite number of purchase states, parameterized by the exhaustion times of the customer's item categories and the transaction time.

The observables of the states correspond to the item categories.  Each item category $i$ is assigned a weight $w_i$ according to the amount of time between the transaction time $T_t$ and $T_{E,i}$ when each item category will be exhausted (based on the usage simulations):

\begin{align} \label{eq:category-weights}
&w_i = - \beta^{-1}_c \exp(-\beta^{-1}_c (T_{E, i}- T_T)) \\
&\beta^{-1}_c \sim U(a,b) \nonumber
\end{align}

where $\beta_c$ the average number of days between when a customer purchases an item and the exhaustion time. The value $\beta_c$ is set separately for each customer by sampling from a uniform distribution. The value $w_i$ is used to model the propensity for a customer to purchase an item they will run out at time $T_{E, i}$ in the future when at the store at time $T_T$.

The item category is chosen by sampling from the item categories with the probability of choosing each item category $i$ proportional to its weight $w_i$.

Markov Models are used to model the customer's purchasing behavior for each item category.  Each state $X_n$ corresponds to an item in that category. The customer's buying habits are determined by the transition probabilities:

\begin{align}
Pr(X_{n+1}&=x|X_n=y) = \nonumber \\
& \left\{ 
  \begin{array}{l l}
   (1.0 - p_l) \, Pr(x,y)  & \quad \text{if $x \neq y$}\\
   p_l & \quad \text{if $x = y$}
  \end{array} \right.
\end{align}

where $w_l$ is the loopback probability, giving the probability of choosing the same item $y$ again. The function $Pr(x,y)$ gives the probability of choosing the item $x$ given that the previous item was $y$:

\begin{equation*}
Pr(x,y) = \frac{\sum_f w_f w_{f,s}(x, y)}{\sum_{j \neq x} \sum_f w_f w_{f,s}(x, j)}
\end{equation*}

where $w_f$ is the weight of a particular field $f$.   The pair of items $x$ and $y$ are weighted  according to the equality of the values for each field:

\begin{equation*}
w_{f,s}(x,y) = \left\{ 
  \begin{array}{l l}
   w_{f,s}  & \quad \text{if $x$.$f$ = $y$.$f$}\\
   1 - w_{f, s} & \quad \text{if $x$.$f$ $\neq$ $y$.$f$}
  \end{array} \right.
\end{equation*}

The weights $w_l$, $w_f$, and $w_{f, s}$ are chosen randomly for each customer:

\begin{align*}
&w_l \sim N(\mu, \sigma^2) \\
&w_f \sim N(\mu, \sigma^2) \\
&w_{f, s} \sim N(\mu, \sigma^2)
\end{align*}

If the weights are greater than 1, they are rounded down to 1.  Likewise, if the weights are less than 0, they are rounded up to 0.

%\textcolor{red}{reorder so that probilities describe habits}

%For example, if a customer tends to buy the same brand of dog food repeatedly, the outgoing edges to all products of the same brand as the product in the current state will be higher than those for other brands' products. If the customer does not care about flavor, the outgoing edges to all products from the current brand will be equal, otherwise the edge weights can be adjusted to prefer a specific combination of brand and flavor.  Likewise, if a customer cares more about flavor or cost, the edge weights will be assigned appropriately.

To simulate an item purchase, the chosen Markov model is transitioned forward by one state. The current states of the Markov models are kept between purchases.  After an item is chosen, the item category's exhaustion time is updated by propagating the item category usage SDE described in Section~\ref{sec:transaction-times}, resulting in an implicit creation of the next possible purchase state in the HMM.

After each purchase, the HMM's current state is transitioned to either the new purchase state or the stop state.  The probability of choosing the next state is

\begin{align*}
&Pr(X_{n+1}=\text{stop}|X_n) = p(\text{stop})\\
&Pr(X_{n+1}=\text{purchase}|X_n) = 1 - p(\text{stop})
\end{align*}

where $p(\text{stop})$ is the probability of choosing the stop state and is given by

\begin{equation*}
 p(\text{stop}) = \frac{w_{\text{stop}}}{w_{\text{stop}} + \sum_i w_i}
\end{equation*}

where $w_{\text{stop}}$ is the weight of the stop state and is a simulation parameter and the weight $w_i$ of each item category $i$ was given earlier in Eq.~\ref{eq:category-weights}.

If the stop state is chosen, the transaction is finished and the next transaction is started according to the procedure presented in Section~\ref{sec:transaction-times}.

%The HMM can be extended to consider additional factors such as the amount of money spent so far in the transaction and the customer's spending habits.

\section{Implementation}
The models and simulations were implemented using Python. The source code is available at \url{https://github.com/rnowling/bigpetstore-data-generator} under the Apache Public License v2.

\section{Evaluation of the Model}
In accordance with our original design goals, we evaluated the model and implementation in terms of their abilities to generate semantically-interesting data and scale from a single local machine to a cluster.

\subsection{Example Analysis of Generated Data}
To evaluate the semantic content of the generated data, we performed two example analyses in support of a fictional advertising campaign using data for 10 stores, 10,000 customers, and five years of transactions generated from the model. The analyses were designed to be realistic and driven by real-world business concerns.  Due to limited advertising budgets, businesses need to decide who to target and what sort of advertisements are likely to be effective for each type of customer.  We analyzed data generated from the model to determine where the advertising campaigns should be targeted geographically and identify customer purchasing profiles.

First, we wanted to identify how close most customers live to a store since advertising to people who live or work too far away from a store is unlikely to be effective and will increase our costs. We analyzed the distribution of distances between customers' homes and their nearest stores.  We found that most customers tend to live within 15 miles of their closest store, suggesting that we should limit our advertising to a 15-mile radius around each our store.

@@

Secondly, we profiled our customers' purchasing habits to optimize the advertising campaign's effectiveness by customizing the advertised products for each customer.  For each customer, we generated a feature vector by computing the frequency with which they would purchase the same brand or flavor from one transaction to the next.  We clustered the customers' feature vectors using the KMeans algorithm with a range of cluster counts. Twenty clusters converged the error.

@@

\subsection{Scaling of Data Size and Run Time}
BigPetStore aims to scale from a local desktop to a large cluster. To evaluate the scaling of the model and implementation, we benchmarked the data generator on a laptop with a 2 GHz Intel Core i7 CPU, 8 GB of RAM, and a 256 GB SSD using the Python implementation with a single thread. Using the test setup, between 1,500 and 2,000 transactions can be generated per second (Table~\ref{tab:benchmarks}). As the customers' transaction simulations are independent of one another, the transaction generation can easily be parallelized so that 1,500-2,000 transactions can be generated per thread per second.

\begin{table}[!t]
%% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{Benchmarks of Data Generator}
\label{tab:benchmarks}
\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
\begin{tabular}{|p{0.75cm}||p{1.2cm}||p{1cm}||p{1.25cm}||p{1cm}||p{1cm}|}
\hline
Stores & Customers & Simulated Time (years) & Transactions & Data Size (MB) & Run Time (min)\\ \hline
10 & 10,000 & 1 & 279,870 & 18 & 3.5 \\ \hline
%10 & 100,000 & 1 & 
100 & 10,000 & 1 & 279,586 & 18 & 3.5 \\ \hline
% 100 & 100,000 & 1 & 
10 & 1,000 & 5 & 123,309 & 8 & 1.1 \\ \hline
100 & 1,000 & 5 & 127,064 & 8 & 1.2 \\ \hline
10 & 10,000 & 5 & 1,275,542 & 84 & 11.0 \\ \hline
%10 & 100,000 & 5 & 
100 & 10,000 & 5 & 1,268,403 & 84 & 10.5 \\ \hline
\end{tabular}
\end{table}

The number of transactions (and hence, data size) grows as $O(N_c T)$ where $N_c$ is the number of customers and $T$ is the amount of time to be simulated. The amount of data generated can be scaled as large as necessary simply by increasing $N_c$ and $T$. Five years of transactions for 100,000 customers will generate approximately 1 GB of data. A terabyte of data can be generated by simulating 100 million customers over five years.

>>>>

% *** SUBFIGURE PACKAGES ***
%\usepackage[tight,footnotesize]{subfigure}
% subfigure.sty was written by Steven Douglas Cochran. This package makes it
% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
% work, it is a good idea to load it with the tight package option to reduce
% the amount of white space around the subfigures. subfigure.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at:
% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
% subfigure.sty has been superceeded by subfig.sty.

\usepackage{caption}
\usepackage{subcaption}

@@

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Adding Local Weather Simulation to the BigPetStore Data Generator for the JVM}

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Ronald J. Nowling}
\IEEEauthorblockA{Red Hat, Inc.\\
Raleigh, NC 27601\\
Email: rnowling@redhat.com}
\and
\IEEEauthorblockN{Jay Vyas}
\IEEEauthorblockA{Red Hat, Inc.\\
Raleigh, NC 27601\\
Email: jvyas@redhat.com}}

@@

\section{Overview of Previous Model}
Our generative model combines various well-known mathematical modeling techniques to simulate the factors affecting customers' purchasing habits.  Each part of the model is derived from \emph{ab initio} assumptions.  In several cases, real data is used to parameterize parts of the model. We described the initial version of the model in \cite{Nowling2014}, which we summarize here.

The model generates data for four types of records: stores, customers, transactions, and transaction items (Figure~\ref{fig:relational-data-model}).  Store records contain both a unique identifier and a location in the form of a 5-digit zip code. Customer records consist of a unique identifier, a name, and a location given as a 5-digit zip code. Transaction records consist of a transaction identifier, a customer identifier, a store identifier, and a time given in days since the beginning of the simulation. Transaction item records contain a transaction identifier and an item description.  The item description is a list of key-value pairs stored as a string.  Key-value pairs are used since item characteristics differ depending on the type of item.  For example, dog food has a brand, a flavor, a size (in lbs), and a per-unit cost, while poop bags have a brand, a color, a count, and a per-unit cost.

@@

\subsection{Stores}
Store records consist of unqiue integer identifiers and locations given as zip codes.  The stores' zip codes are sampled from a probability mass function (PMF) where probabilities are linearly proportional to the zip code's population and exponentialy proportional to the zip code's median household income. The population and household income data for the zip codes were taken from from the U.S. Census American Community Survey \cite{ACS}.

\subsection{Customers}
Customer records consist of uniquer integer identifers, first names, last names, and locations given as zipcodes.  Names are generated using data from the Name Database\cite{NameDB}. Each record in the database gives a name, a weight, and flag indicatings if the name can be used as a first name, a last name, or both.  PMFs generated for the first and last names using the weights.  The customer's name is generated by sampling a first name and a last name from each PMF respectively.  Zip codes are sampled from a PMF where the probabilities are exponentially distributed according to the distance (in miles) to the nearest store's location. Distances are calculated using latitude and longitude coordinates taken from the Zip Code Database Project \cite{Zips}).

\subsection{Transactions}
Transactions records consist of customer, transaction, and store identifiers, the date and time of the transaction (as a real number indicating the number of days since the start of the simulation), and a list of purchased products. The customer and transaction identifiers form a unique composite key for each transaction record.

Transactions are simulated individually for each customer.  Transaction times are sampled using a Monte Carlo process.  Transaction times are proposed by sampling offsets before the exhaustion of products from an exponential distribution, with the average time parameterized separately for each customer by sampling from a uniform interval.  Acceptance probabilities are given by a separate PDF which can be used to model the effect of with factors such as store hours and customer work schedules.  Currently, the acceptance PDF only ensures that a new transaction time occurs after the previous transaction.  The proposed transaction time is accepted if a random number sampled uniformally from $[0, 1)$ is less than the acceptance probability. Otherwise, a new transaction time is proposed.

Product purchases are simulated by a separate process.  First, a state is chosen from the set of product categories and a stop state which indicates the end of the transaction.  Each product category's weight is exponentially proportional to the elapsed time between the transaction and exhaustion of the product category. The weight of the stop state is given by a constant which is zero until the first product is chosen.

Once a product category has been chosen, the process of choosing a product within the category is modelled by a Markov Model.  \textcolor{red}{The transition probabilites are determined by ...}. A product is chosen by advancing the state of the model by simulate one transition.  The state of the Markov Model is preserved between transactions.

Once a product has been chosen, the customer's inventory is updated and the exhaustion time is determined by simulating the usage of the products by numerically integrating a stochastic differential equation (SDE).  The purchasing process is restarted, using the updated exhaustion time.  If the stop state is chosen, the next transaction is generated.

%\begin{figure}[!t]
%  \centering
%  \includegraphics[width=3.5in]{figures/transaction_time_model.eps}
%  \caption{Transaction Time Monte Carlo Simulation}
%  \label{fig:trans_time}
%\end{figure}

%\begin{figure}[!t]
%  \centering
%  \includegraphics[width=3.5in]{figures/transaction_purchases_model.eps}
%  \caption{Transaction Purchases Monte Carlo Simulation}
%  \label{fig:purchases_sim}
%\end{figure}

\begin{figure}[!t]
  \centering
  \includegraphics[width=3in]{figures/dog_food_markov_model.eps}
  \caption{Example Markov Model for Dog Food}
  \label{fig:dog_food_markov_model}
\end{figure}

\begin{figure}[!t]
  \centering
  \includegraphics[width=3.5in]{figures/item_usage_simulation.pdf}
  \caption{Example Product Usage Simulation}
  \label{fig:product_usage_sim}
\end{figure}

\section{Improvements to the Data Generator}
\subsection{Products}
The original data generator relied on a hand-generated JSON document containing product descriptions.  Due to the labor requirements, only a handful of products were available in each category with only 3-4 fields per category. To increase the number and variety of purchasing patterns in the model, a product enumerator was implemented.  Given a set of fields and possible values, the enumerator generates all combinations of products.

Rules are used to filter out unwanted combinations of field values.  For example, the fictional brand of dog food ``Chef Corgi'' might be targeted towards budget-oriented consumers.  As such, the brands' products might exclude expensive ingredients and marketing (salmon, organic ingredients, etc.).  Products that do not satisfy the filters are removed from the list of possible products.

Using the product enumerator, the number of products across four categories has been increased from \textcolor{red}{40 to over 3,000, and the number of fields has increased from 3-4 to upwards of 10 per category}.

Optional products were introduced.  Examples of optional products include snacks, collars, sweaters, and toys. Exhaustion of optional products does not trigger transactions.  Said another way, the optional products' exhaustion times are not used when proposing transaction times.  There is no distinction between optional and mandatory products when simulating purchases within a transacton.

\textcolor{red}{weather-triggered or seasonal products. product PDFs?}

\subsection{Customer and Transaction Stores}
In the data generator prototype, the store for each each transaction was picked uniformally from the set of all stores.  To improve realism and support the incorporation of geospatial patterns, each customer is now associated with a specific store and all of a customer's transactions occur at that store.

Customers' associated stores are sampled from a PMF with weights proportionally to population of each stores' zip code.  The model for sampling the customers' locations was updated so that each zip code $z$ is weighted exponententially according to the distance from the zip code $s$ of the customers' associated store:

\begin{equation}
w(s, z) = \lambda \exp(-\lambda d(s, z))
\end{equation}

where $\lambda$ is the inverse desired average distance.

\subsection{Weather Model}

\textcolor{red}{TODO why?}

Weather data is generated using a local weather model -- i.e., weather is not explicitly correlated between regions -- so that each region can be simulated independently for computational efficiency.  The model was derived and evaluated by comparing to local weather data downloaded from the NOAA QCLCD \textcolor{red}{cite}.  The goal of the model was not to accurately predict day-to-day weather patterns but to reproduce temporal and static statistical properties.  The model is similar other models available in the literature \textcolor{red}{cite}.

The models were derived for daily average temperatures ($T(t)$, $^\circ$F), daily total precipitation ($P(t)$, in), and daily average wind speed ($V(t)$, mph). The temperature for each location is modeled as a first-order Fourier series plus noise generated by an Ornstein-Uhlenbeck process $Z(t)$ \textcolor{red}{cite}:

\begin{align}
&T(t) = a_0 + a_1\sin\Big(\frac{-2 \pi t}{365}\Big) + a_2\cos\Big(\frac{-2 \pi t}{365}\Big) + Z(t) \\
&dZ(t) = - \gamma Z(t) dt + \sigma dW_t
\end{align}

The stochastic differential equation (SDE) for the Ornstein-Uhlenbeck process is integrated numerically using the Euler-Maruyama method \cite{Klouden13}:

\begin{align}
&Z_{t+1} = -\gamma Z_t \Delta t + \sigma \sqrt{\Delta t} \Delta W_{t+1} \\
&Z_0 = 0 \nonumber
\end{align}

The daily total precipitation for each region is assumed to be independently distributed from day-to-day and sampled from an exponential distribution:

\begin{equation}
P(t) \sim Exp(\lambda)
\end{equation}

The daily average wind speed for each region is modeled using a first-order Fourier series and stochastic equation:

\begin{align}
&V(t) = b_0 + b_1\sin\Big(\frac{-2 \pi t}{365}\Big) + b_2\cos\Big(\frac{-2 \pi t}{365}\Big) + X(t) \\
&X(t) \sim Erlang(k, \theta)
\end{align}

The average temperature $a_0$, $a_1$, $a_2$, inverse daily average precipitation $\lambda$, average daily wind speed $b_0$, $b_1$, $b_2$, $k$, and $\theta$ are \textcolor{red}{fit to the observed data (how?)} from NOAA QCLCD for each zip code.  The damping coefficient $\gamma$ is determined empirically.

Three years of observed daily average temperature, total precipitation, and average wind speed from the NOAA QCLCD data for South Bend, IN from October 2011 to September 2014 are shown in Figure~\ref{fig:weather_model} alongside simulated data from the model.  The simulated data is able to produce temporal and stationary patterns similar to those of the real data.

From the temperature, precipitation and wind speed, the snowfall ($S(t)$, in), rainfall ($R(T)$, in), and wind chill ($C(T)$, $^\circ$F) \textcolor{red}{cite} are calculated as follows:

\begin{align}
&r(t) = \frac{1}{1 + \exp(-a (T - b))} \\
&S(t) = 10 (1 - r(t)) P(T) \\
&R(T) = r(t) P(T)
\end{align}

\begin{align}
C(T) = 35.74 + &0.6215T(t) - 35.75V^{0.16}(t) + \\ \nonumber
&0.4275T(t)V^{0.16}(t)
\end{align}

The snowfall, rainfall, and wind chill are used as input to a conditional probability density function (PDF) giving the probability of a transaction occuring on the date of time $t$ given the weather conditions:

\begin{equation}
P(\text{trans}|t, S(T), R(T), C(T)) = \dots
\end{equation}

\begin{figure*}[!t]
  \centering
  \begin{subfigure}[b]{0.4\textwidth}
    \includegraphics[width=\textwidth]{figures/sim_temp.pdf}
    \caption{Daily Temperature}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.4\textwidth}
    \includegraphics[width=\textwidth]{figures/sim_temp_hist.pdf}
    \caption{Temperature Histogram}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.4\textwidth}
    \includegraphics[width=\textwidth]{figures/daily_precip.pdf}
    \caption{Daily Precipitation}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.4\textwidth}
    \includegraphics[width=\textwidth]{figures/daily_precip_hist.pdf}
    \caption{Precipitation Histogram}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.4\textwidth}
    \includegraphics[width=\textwidth]{figures/daily_wind_speeds.pdf}
    \caption{Daily Wind Speed}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.4\textwidth}
    \includegraphics[width=\textwidth]{figures/daily_wind_speed_hist.pdf}
    \caption{Wind Speed Histogram}
  \end{subfigure}

  \caption{NOAA QCLCD weather data alongside simulated data for South Bend, IN for October 2011 to September 2014.}
  \label{fig:weather_model}  
\end{figure*}

\subsection{Implementation}
The data generator was previously prototyped in Python.  Once the utility of the generator was validated, a new production implementation of the data generator was implemented and released as a Java library. Java was chosen for compatibility with popular big data processing systems such as Apache Hadoop \textcolor{red}{cite} and Apache Spark \textcolor{red}{cite} and other JVM languages such as Scala and Clojure.  The Java implementation has ~5,500 lines of code, 110 classes, and 90 unit tests.

Users interact with the library through six primary classes (\texttt{DataLoader}, \texttt{StoreGenerator}, \texttt{CustomerGenerator}, \texttt{ProductGenerator}, \texttt{PurchasingProfileGenerator}, and \texttt{TransactionGenerator}) and five data model POJOs (\texttt{InputData}, \texttt{Store}, \texttt{Customer}, \texttt{Product}, and \texttt{Transaction}).  Broad support for multiple parallelization paradigms was a primary design requirement.  Generators interact only through standard Java collections containing serializable POJOs and are designed to support multiple instantiated copies to support parallelization.  As such, parallelization is as simple as creating as many independent instances as needed and combining the resulting Java collections before each stage. An example driver implemented with Apache Spark with support for parallel and distributed computing is presented in Section~\ref{sec:spark}.  For development purposes, the library includes a single-threaded command-line driver.

\begin{figure}[!t]
  \centering
  \includegraphics[width=3.5in]{figures/bps_flowchart.eps}
  \caption{BigPetStore Flowchart}
  \label{fig:bps_flowchart}
\end{figure}

The source code is available at \url{https://github.com/rnowling/bigpetstore-data-generator} under the Apache Public License v2. Jars are published via \textcolor{red}{\url{bintray}} for easy integration with build systems that support automatic dependency resolution.

\section{Parallelization with Spark Driver} \label{sec:spark}
\textcolor{red}{TODO code listing}

\section{Evaluation of the Model}
To evaluate the semantic content of the generated data, we performed two example analyses in support of a fictional advertising campaign using data for 10 stores, 10,000 customers, and five years of transactions generated from the model. The analyses were designed to be realistic and driven by real-world business concerns.  Due to limited advertising budgets, businesses need to decide who to target and what sort of advertisements are likely to be effective for each type of customer.  We analyzed data generated from the model to determine where the advertising campaigns should be targeted geographically and identify customer purchasing profiles.

\subsection{Weekly Transaction Counts by Store}
First, we wanted to identify how close most customers live to a store since advertising to people who live or work too far away from a store is unlikely to be effective and will increase our costs. We analyzed the distribution of distances between customers' homes and their nearest stores.  We found that most customers tend to live within 15 miles of their closest store, suggesting that we should limit our advertising to a 15-mile radius around each our store.

@@

\subsection{Customer Segmentation}
Secondly, we profiled our customers' purchasing habits to optimize the advertising campaign's effectiveness by customizing the advertised products for each customer.  For each customer, we generated a feature vector by computing the frequency with which they would purchase the same brand or flavor from one transaction to the next.  We clustered the customers' feature vectors using the KMeans algorithm with a range of cluster counts. Twenty clusters converged the error.

@@

\textcolor{red}{compare new and old}

\subsection{Product Recommendations}

\textcolor{red}{Need to see if we can evaluate by making vectors sparser}

\subsection{Impact of Weather on Shopping Patterns by Region}

\textcolor{red}{TODO: effect of weather, variations by region}

\section{Discussion and Conclusion}
The BigPetStore data generator is a domain-driven mathematical model and simulation for generating data with rich patterns.  We described several updates to the model designed to increase the number and variety of correlations.  By implementing a product enumerator, we increased the number of products and product fields, resulting in a significant increase in the number of possible purchasing profiles.  The addition of a model for local weather simulation enabled the incorporation of geospatial and additional temporal correlations.  A new JVM library implementation was described.  Support for parallelization and interopability with tools in the big data ecosystem was demonstrated using an Apache Spark driver.

\textcolor{red}{The math model was validated using ...}

We see multiple opportunities for building on the work presented.  \textcolor{red}{...}

We intend to expand the model to incorporate additional factors.  Deeper integration of existing fields such as location and purchasing profiles (e.g., to model regional purchasing preferences) will increase the variety of the semantic information encoded in the generated data. The incorporation of weather and climate data can be used to influence when customers shop, the types of products they buy, and the amount of products purchased. For example, customers would be less likely to shop during snow storms, more likely to buy items such as winter apparel for their pets, and more likely to purchase bulk quantities to reduce the number of transactions. Modeling of time-dependent events such as sales, evolution of customer purchasing profiles over time, and ``life events,'' such as the birth or passing of pets, will enable more interesting time-series analysis of the data.  We would also like to expand the scope of the model to incorporate business processes (such as inventory management, customer complaints, employees, etc.), thus enabling queries about the relationship between internal business process and customer behavior.
